---
title: "Final CMSC320 Project"
output: html_document
---


### Welcome to our tutorial
Written by Sam Jalowicz and Nick Kibbey

# Part 0: Purpose

### What is this project?
This tutorial is written to walk you through the typical data science pipeline.

### Who is this written for?
You! Or anybody interested in data science. This tutorial is written for a beginner level. 

### Requirements
Novice coding experience helps and we assume you have been able to use R and have an editor you are comfortable with. This project is written in R 3.6.0, RMarkdown using RStudio 1.1.463. 


We will be making use of tidyverse among other packages and if you do not yet know what tidyverse is I personally recommend you to get aquainted with it at https://www.rdocumentation.org/packages/tidyverse/versions/1.2.1

### What will you be doing?
We're not sure

# Part 1

### Setup
We are using tidyverse and when it loads there will probably be warnings and other confusing information that may pop up but there is nothing to worry about if the proceeding code works fine.

```{r setup1, include=FALSE}
library(tidyverse)
```
```{r setup2, eval=FALSE}
library(tidyverse)
```

### Download the data 
Download from https://wri-public-data.s3.amazonaws.com/ImpactHack/Refugees.zip
and unzip this folder, I left mine in my downloads but if you put the "aslyum_by_country_origin2.csv" in the same folder as your R file then you won't need the prefix that I used for the my_refugee_addr variable.

This data will include all the countries the U.S. state department cares to include in their dataset and the amount of refugees from that country by year from 2000 - 2017.

```{r getRefugees}
# CHANGE HERE to where you downloaded your file
# my_refugee_addr <- "/Users/sjalowicz/Desktop/Refugees/aslyum_by_country_origin2.csv"
my_refugee_addr <- "Refugees/aslyum_by_country_origin2.csv"

# Variable that stores the data
refugee_tab <- read_csv(my_refugee_addr)
# Displaying a preview of the data stored in our R program
refugee_tab
```


### Data Management

Our objective is to be able to make a graph that has time on the X axis and number of refugees on the Y axis but that is difficult with our current data because the year is an attribute and not a value. Our data is too wide and we want it more skinny so that the specific year is a value we can plot and to do that across 17 years we will have 17 entries for each country instead of 1 country with 17 attributes relating to that year's refugees. 

##### TL;DR, We need to modify our table if we want to graph over time

#### How are we doing that?
By using my lazy script to write the R code that creates 17 temp variables that will be merged into one data frame. The whole R code for this task is in the original markdown and the lazy script I used to help me is aptly named lazy.sh.

What you see in the next snippet is the basic gist of what was repeated 17 times with minor adjustments.

```{r dataMod1, include=FALSE}
# Please forgive me because this is about to get tedious and R is a functional language
# Which means without macros the pipeline I chose is not the best but it will work
# Again I am sorry but sometimes data science makes you do mundane parsing 
# You also learn that the smarter way takes longer to do so the stupid way is used more often
temp2000 <- refugee_tab %>%
select(name, yr_2000) %>%
mutate(num_refugees=yr_2000, year=2000)
temp2000$yr_2000 <- NULL

temp2001 <- refugee_tab %>%
select(name, yr_2001) %>%
mutate(num_refugees=yr_2001, year=2001)
temp2001$yr_2001 <- NULL

temp2002 <- refugee_tab %>%
select(name, yr_2002) %>%
mutate(num_refugees=yr_2002, year=2002)
temp2002$yr_2002 <- NULL

temp2003 <- refugee_tab %>%
select(name, yr_2003) %>%
mutate(num_refugees=yr_2003, year=2003)
temp2003$yr_2003 <- NULL

temp2004 <- refugee_tab %>%
select(name, yr_2004) %>%
mutate(num_refugees=yr_2004, year=2004)
temp2004$yr_2004 <- NULL

temp2005 <- refugee_tab %>%
select(name, yr_2005) %>%
mutate(num_refugees=yr_2005, year=2005)
temp2005$yr_2005 <- NULL

temp2006 <- refugee_tab %>%
select(name, yr_2006) %>%
mutate(num_refugees=yr_2006, year=2006)
temp2006$yr_2006 <- NULL

temp2007 <- refugee_tab %>%
select(name, yr_2007) %>%
mutate(num_refugees=yr_2007, year=2007)
temp2007$yr_2007 <- NULL

temp2008 <- refugee_tab %>%
select(name, yr_2008) %>%
mutate(num_refugees=yr_2008, year=2008)
temp2008$yr_2008 <- NULL

temp2009 <- refugee_tab %>%
select(name, yr_2009) %>%
mutate(num_refugees=yr_2009, year=2009)
temp2009$yr_2009 <- NULL

temp2010 <- refugee_tab %>%
select(name, yr_2010) %>%
mutate(num_refugees=yr_2010, year=2010)
temp2010$yr_2010 <- NULL

temp2011 <- refugee_tab %>%
select(name, yr_2011) %>%
mutate(num_refugees=yr_2011, year=2011)
temp2011$yr_2011 <- NULL

temp2012 <- refugee_tab %>%
select(name, yr_2012) %>%
mutate(num_refugees=yr_2012, year=2012)
temp2012$yr_2012 <- NULL

temp2013 <- refugee_tab %>%
select(name, yr_2013) %>%
mutate(num_refugees=yr_2013, year=2013)
temp2013$yr_2013 <- NULL

temp2014 <- refugee_tab %>%
select(name, yr_2014) %>%
mutate(num_refugees=yr_2014, year=2014)
temp2014$yr_2014 <- NULL

temp2015 <- refugee_tab %>%
select(name, yr_2015) %>%
mutate(num_refugees=yr_2015, year=2015)
temp2015$yr_2015 <- NULL

temp2016 <- refugee_tab %>%
select(name, yr_2016) %>%
mutate(num_refugees=yr_2016, year=2016)
temp2016$yr_2016 <- NULL

temp2017 <- refugee_tab %>%
select(name, yr_2017) %>%
mutate(num_refugees=yr_2017, year=2017)
temp2017$yr_2017 <- NULL
```


```{r dataMod2, eval=FALSE}
# Please forgive me because this is about to get tedious and R is a functional language
# Which means without macros the pipeline I chose is not the best but it will work
# Again I am sorry but sometimes data science makes you do mundane parsing 
# You also learn that the smarter way takes longer to do so the stupid way is used more often
temp2000 <- refugee_tab %>%
  select(name, yr_2000) %>%
  mutate(num_refugees=yr_2000, year=2000)
temp2000$yr_2000 <- NULL

```

This is the code that will bind all our temp variables together into one tall data frame.

```{r dataMod3}
refugee_df <- bind_rows(temp2017, temp2016, temp2015, temp2014, temp2013, temp2012, temp2011,
                        temp2010, temp2009, temp2008, temp2007, temp2006, temp2005, temp2004,
                        temp2003, temp2002, temp2001, temp2000)
refugee_df
```

Now we will load in our population set. We only care about the year and population columns, so we remove the rest. These columns are initially set to chr attributes, so in order to work with them later on, we need to convert them. We type convert the year column to an integer, and we type convert the population column to a double. In order to convert the population column, we must first remove the commas from the numbers. Now we filter the data so that it includes only information from 2000-2017, as that is what our refugee data is looking at.

```{r echo=TRUE}
population_df <- read_csv("/Users/sjalowicz/Desktop/World Population.csv")
population_df$population <- as.numeric(gsub(",","",population_df$population))
population_df <- population_df %>%
  group_by(year) %>%
  select(year, population) %>%
  type_convert(cols(year = col_integer())) %>%
  type_convert(cols(population = col_double())) %>%
  filter(year >= 2000, year <= 2017)
population_df
```


Now isn't that better?

# Part 2: Exploratory data analysis

In order to get a better sense of the data we're working with, we perform exploratory data analysis. We try to manipulate the data to see if there are trends that we can work with, or see if there any correlations we can identify.

Here we decided to create a map of refugee data around the world, to get a better picture of the distribution of refugee populations. The areas in yellow have lower numbers of refugees, while the areas in orange have more, and red has the highest. We did this by utilizing the rworldmap library. From this map you can see areas of interest in 2010, and can try to research what was going on in specific countries in 2010 that may have led to these statistics.

```{r chloroplast}
library(rworldmap)
refugees2010 <- refugee_df %>%
  filter(year==2010) 
  
  
ref_map <- joinCountryData2Map(refugees2010, joinCode = "NAME", nameJoinColumn = "name")

#plot the map
mapCountryData(ref_map, nameColumnToPlot='num_refugees') 
```

From here we wanted to see if there was a realtionship between time and the total number of refugees. We summed up the number of refugees in each country based on the year, and created a new dataframe that contained this information. We did not want to overwrite the original data frame, as we will use this again later on. We then plotted the number of refugees over time, and added a regression line in order to see the general trend over time.

```{r echo=TRUE}
library(tidyverse)
library(dplyr)

ref_totals <- refugee_df %>%
  group_by(year) %>%
  select(year, num_refugees) %>%
  summarize(total=sum(num_refugees, na.rm = TRUE))


ref_totals %>%
  ggplot(mapping=aes(x=year, y=total)) +
    geom_smooth(method=lm) +
    geom_point() + labs(title="Number of Total Refugees Over Time",
          x="Year", y="Number of Refugees")

```

We also wanteed to see if there was a correlation between world population and time. We plotted the total population for each year, and added a regression line to this graph as well. As you can see here, this data appears to be more consistent over time than the refugee data that we just plotted.

```{r echo=TRUE}
population_df %>%
  filter(year >= 2000,year <= 2017) %>%
  ggplot(mapping=aes(x=year, y=population)) +
    geom_smooth(method=lm) +
    geom_point() + labs(title="Total World Population Over Time",
          x="Year", y="Population")
```

We also want to explore trends in individual countries over time. From looking at the initial data, we saw a decrease in Angola refugees, and wanted to explore that. We plotted the data, and added a regression line.

```{r echo=TRUE}
refugee_df %>%
  filter(name == "Angola") %>%
  ggplot(mapping=aes(x=year, y=num_refugees)) +
    geom_smooth(method=lm) +
    geom_point() + labs(title="Number of Refugees in Angola Over Time",
          x="Year", y="Number of Refugees")
```

We chose to go a step further and break it down into four distinct time periods. This can be useful when trying to figure out the underlying causes in the refugee decrease. You can see the steepest decrease in the 2000-2004 time period, which may be due to the fact that id when Angolas 30 year civil war was coming to an end, and the Angolan govenment began assisting those that were displaced.

Source: https://www.unhcr.org/en-us/news/latest/2007/3/4607b7d24/repatriation-angola-officially-ends-410000-refugees-home.html

```{r echo=TRUE}
library(tidyverse)

refugee_df %>%
  filter(name == "Angola") %>%
  mutate(cut_yearID = cut(x=year,breaks=4)) %>%
  group_by(cut_yearID, name) %>%
  rowid_to_column() %>%
  ggplot(mapping=aes(x=year, y=num_refugees)) +  
    facet_wrap(~cut_yearID) +
    geom_point() +
    geom_smooth(method=lm)
```

From here we wanted to look at a country that saw an increase in refugees, and try to figure out why this may be the case. We chose to plot the data for Syria and added a regression line.

```{r echo=TRUE}
refugee_df %>%
  filter(name == "Syria") %>%
  ggplot(mapping=aes(x=year, y=num_refugees)) +
    geom_smooth(method=lm) +
    geom_point() + labs(title="Number of Refugees in Syria Over Time",
          x="Year", y="Number of Refugees")
```

We broke this one down into four parts as well, and the results did not disappoint. You can see relatively consistent numbers in the first two graphs, which contain data from 2000-2008. In the third one you can see that in 2012 there was a steep jump in the number of refugees, which gives that regression line a greater slope. In the fourth graph you see a steady increase over the years, which is why the initial graph above shows an increase in the number of refugees in Syria over time. Upon further research, 2011/2012 is when the Syrian Civil War began. As with the Angola data, we are starting to see a pattern. It's starting to look like many of these refugees can be attributed to times of civil war and distress in individual countries. 

Source: https://www.worldvision.org/refugees-news-stories/syrian-refugee-crisis-facts

```{r echo=TRUE}
library(tidyverse)
#library(nycflights13)

refugee_df %>%
  filter(name == "Syria") %>%
  mutate(cut_yearID = cut(x=year,breaks=4)) %>%
  group_by(cut_yearID, name) %>%
  rowid_to_column() %>%
  ggplot(mapping=aes(x=year, y=num_refugees)) +  
    facet_wrap(~cut_yearID) +
    geom_point() +
    geom_smooth(method=lm)
```

In order to put it in perspective, we wanted to plot the trends on the same graph. While Angola may look like it has a steep decrease on its own, it looks relatively consistet on a larger scale when compared to the increase in Syrian refugees.

```{r echo=TRUE}
refugee_df %>%
  filter(name == "Angola" | name == "Syria") %>%
  ggplot(mapping=aes(x=year, y=num_refugees, color=name)) +
    geom_smooth(method=lm) +
    geom_point() + labs(title="Number of Refugees in Select Countries Over Time",
          x="Year", y="Number of Refugees")
```

# Part 3: Hypothesis Testing and Machine Learning

In order to figure out how "good" our data is, we perform hypothesis testing. We create a hypothesis and perform different tests to see how confident our results are. You can either reject the hypothesis or not reject the hypothesis. In our case we are going to compare the total refugee population with the total world population. 

Our Hypothesis: World refugee rates and world population growth are proportional.

We will use linear regression to get a sense of the linear relationship between time and refugee population, and a sense of the linear relationship between time and total world population.

We will then use anova to compare the two relationships, and see if one rate is increasing faster than the other, thus proving our hypothesis wrong.

First we use lm to analyze the refugee data.

```{r echo=TRUE}
ref_lm <- lm(year~total, data=ref_totals)
summary(ref_lm)
```

```{r echo=TRUE}
coef(ref_lm)
```

Next we use lm to get the total population data.

```{r echo=TRUE}
pop_lm <- lm(year~population, data=population_df)
summary(pop_lm)
```

```{r echo=TRUE}
coef(pop_lm)
```

We use anova to compare the two data sets.

```{r echo=TRUE}
anova(ref_lm,pop_lm)
```

From here we can conclude that the total refugee population is increasing at a greater rate (3.074632e-07) than the total population rate (1.246577e-08), which means our hypothesis is proved wrong, as we predicted they would be proportional. As we hinted at before, this can be attributed to many underlying factors going on within a country. In the case of Angola and Syria specifically, the respective countries civil wars played a huge role in refugee population variation. The total population apears to be steadily increasing over time, and does not seem to be reliant on individual countries and events that are taking place.





